Maximum likelihood estimation of cardinalities\\ from HyperLogLog sketches

### 2016-04-25 paper/paper.tex
some minor changes

The number of bits of the consumed hash value $\symPrecision+\symRegRange$ limit the maximum cardinality that can be tracked. Obviously, if the cardinality reaches values in the order of $2^{\symPrecision+\symRegRange}$, hash collisions will become more apparent  , which will reduce estimation accuracy drastically.

@@

For large cardinalities in the order of $2^{\symPrecision+\symRegRange}$ for which a lot of registers already are already in a saturated state meaning that they have already reached the maximum possible value $\symRegRange + 1$, the \emph{raw} estimate underestimates the cardinalities. For the 32-bit hash value case $(\symPrecision+\symRegRange=32)$ which was considered in \cite{Flajolet2007}, following correction formula was proposed to take these saturated registers into account
\begin{equation}
\label{equ:large_range_estimate}
\symCardinalityEstimate_\text{large}
=
-2^{32}\log\left(1-\frac{\symCardinalityOriginalEstimate}{2^{32}}\right).
\end{equation}

The entire original estimation algorithm as presented in \cite{Flajolet2007} including small and large range corrections is summarized by \algoref{alg:estimate_original}. 
\begin{algorithm}
\caption{Original procedure for estimating the cardinality from a  HyperLogLog data structure using 32-bit hash values ($\symPrecision+\symRegRange = 32$) for insertion of data items \cite{Flajolet2007}.}
\label{alg:estimate_original}
\begin{algorithmic}
\Function {EstimateCardinality}{$\vec{\symRegVal}$}
\State $\symNumReg \gets \dim \vec{\symRegVal}$
\State $\symCardinalityOriginalEstimate = \symAlpha_\symNumReg \symNumReg^2\left(\sum_{\symIndexJ=1}^\symNumReg 2^{-\symRegVal_\symIndexJ}\right)^{-1}$
\Comment \emph{raw} estimate \eqref{equ:intermediate_estimate}
\If{$\symCardinalityOriginalEstimate\leq \frac{5}{2}\symNumReg$}
\State $\symCount_0 = \left|\{\symIndexJ |\symRegVal_\symIndexJ=0\}\right|$
\If{$\symCount_0\neq0$}
\State \Return $\symNumReg \log(\symNumReg / \symCount_0)$ \Comment small range correction \eqref{equ:small_range_estimate}
\Else
\State \Return $\symCardinalityOriginalEstimate$
\EndIf
\ElsIf{$\symCardinalityOriginalEstimate\leq \frac{1}{30}2^{32}$}
\State \Return $\symCardinalityOriginalEstimate$
\Else
\State\Return $-2^{32}\log(1-\symCardinalityOriginalEstimate/2^{32})$\Comment large range correction \eqref{equ:large_range_estimate}
\EndIf
\EndFunction
\end{algorithmic}
\end{algorithm}
The relative estimation error for the original method is shown in \autoref{fig:original_estimate}. Unfortunatley, as can be clearly seen, the ranges where the estimation error is small for $\symCardinalityOriginalEstimate$ and $\symCardinalityEstimate_\text{small}$ do not overlap. Therefore, the estimation error is much larger near the transition region. To reduce the estimation error for cardinalities close to this region, it was proposed to correct $\symCardinalityOriginalEstimate$ for bias. The empirically determined bias correction data can be either stored as set of interpolation points \cite{Heule2013}, as lookup table \cite{Rhodes2015}, or as best-fitting polynomial \cite{Sanfilippo2014}.

@@

The large range correction is not satisfying either. The large range correction does not reduce the estimation error. It even increases the estimation error. However, instead of underestimating the cardinalities, they are now overestimated. A simple approach to avoid the large range correction at all is the use of larger hash values $(\symPrecision+\symRegRange>32)$ which shift the biased region of $\symCardinalityOriginalEstimate$ to higher cardinalities. However, in case $\symRegRange \geq 30$ 6-bit registers as used in \cite{Heule2013} are required to allow representing all values in range $[0, \symRegRange+1]$. If 64-bit hash values are used as proposed in \cite{Heule2013} $(\symPrecision+\symRegRange=64)$, a correction for large cardinalities is not required. Cardinalities in the order of $2^{64}$ for which saturated register become a problem never occur in practice.

@@

\section{Cardinality estimation}
\label{sec:cardinality_estimation}
In the following we describe a new cardinality estimation method that does not require special handling for small and large cardinalities and which also does not require any empirically determined constants or corrections. The method is based on the maximum likelihood method.

The maximum likelihood method was already applied to HyperLogLog sketches in \cite{Clifford2012}. However, there, a different element insertion algorithm was assumed that works with time complexity $\symBigO(\symNumReg)$, because each register calculates its own hash value from the inserted element and updates its current value accordingly. In comparison, \algoref{alg:insert} uses only a single hash value that defines which register needs to be updated. As a consequence, the register values are dependent, which makes the joint probability distribution of all register values more complex.

In the following we use the Poisson approximation in order to simplify the model. Despite this approximation the maximum likelihood estimate will give us high quality estimates for the entire cardinality range as simulation results will show later. Furthermore, we will demonstrate that consequent application of the maximum likelihood method to the Poisson model directly reveals that the cardinality estimate needs to be roughly proportional to the harmonic mean for intermediate cardinality values. The history of the HyperLogLog algorithm shows that the \emph{raw} estimate \eqref{alg:estimate_original} was first found after several attempts using the geometric mean \cite{Durand2003, Flajolet2007}.

@@

Under the assumption of an ideal hash function the probability mass function for the register values $\vec{\symRegVal}=(\symRegVal_1,\ldots,\symRegVal_\symNumReg)$ of a HyperLogLog sketch with parameters $\symPrecision$ and $\symRegRange$ is given by
\begin{equation}
\label{equ:multinomialProbabilityMass}
\symProbabilityMass(\vec{\symRegVal}\vert\symCardinality)
=
\sum_{\symCardinality_1+\ldots+\symCardinality_\symNumReg = \symCardinality} \binom{\symCardinality}{\symCardinality_1,\ldots\symCardinality_\symNumReg}
\frac{1}{\symNumReg^\symCardinality}\prod_{\symIndexJ=1}^\symNumReg \symRegProbability_{\symCardinality_\symIndexJ\symRegVal_\symIndexJ},
\end{equation}
where $\symCardinality$ is the cardinality. The $\symCardinality$ distinct elements are distributed over all $\symNumReg$ registers according to a multinomial distribution with equal probabilities. $\symRegProbability_{\symNu\symKappa}$ is the probability that a register is equal to $\symKappa$ after inserting $\symNu$ distinct elements
\begin{equation}
\symRegProbability_{\symNu\symKappa} 
:=
\begin{cases}
1 & \symNu=0 \wedge \symKappa = 0\\
0 & \symNu=0 \wedge 1\leq\symKappa\leq\symRegRange+1\\
0 & \symNu\geq1 \wedge \symKappa = 0\\
\left(1-\frac{1}{2^\symKappa}\right)^\symNu - \left(1-\frac{1}{2^{\symKappa-1}}\right)^\symNu & \symNu \geq 1 \wedge 1\leq\symKappa\leq\symRegRange\\
1 - \left(1-\frac{1}{2^{\symRegRange}}\right)^\symNu & \symNu\geq 1 \wedge \symKappa = \symRegRange +1 \\
\end{cases}
\end{equation}
Note that the order of register values $\symRegVal_1,\ldots,\symRegVal_\symNumReg$ is not important for the estimation of the cardinality. More formally, the multiset $\lbrace\symRegVal_1,\ldots,\symRegVal_\symNumReg\rbrace$ is a sufficient statistic for $\symCardinality$.
Since the values of the multiset are all in the range $[0, \symRegRange+1]$ the multiset can also be represented as $\lbrace\symRegVal_1,\ldots,\symRegVal_\symNumReg\rbrace = 0^{\symCount_0}1^{\symCount_1}\cdots\symRegRange^{\symCount_{\symRegRange}}(\symRegRange+1)^{\symCount_{\symRegRange+1}}$ where $\symCount_\symIndexJ$ is the multiplicity of value $\symIndexJ$. As a consequence, the multiplicity vector $\vec{\symCount} := (\symCount_0,\ldots,\symCount_{\symRegRange+1})$ is also a sufficient statistic for the cardinality. In addition, this vector also contains all the information about the HyperLogLog sketch that is required for  cardinality estimation: $\symRegRange=\dim(\vec{\symCount})-2$ and  $\symPrecision = \log_2 \left\Vert\vec{\symCount}\right\Vert_1$.

@@

As supported by simulations that will be presented later, this approximation works well for the entire range of cardinalities. In order to get an estimate for $\symPoissonRateEstimate$, maximum-likelihood estimation is used which works out quite well under the Poisson model. The log-likelihood and its derivative are given by
\begin{equation}
\label{equ:log_likelihood_single}
\log \mathcal{\symLikelihood}(\symPoissonRate\vert\vec{\symRegVal}) = 
-\frac{\symPoissonRate}{\symNumReg}\sum_{\symRegVal=0}^{\symRegRange}\frac{\symCount_\symRegVal}{2^\symRegVal}+ 
\sum_{\symRegVal=1}^{\symRegRange}\symCount_\symRegVal \log\left(1-e^{-\frac{\symPoissonRate}{\symNumReg 2^\symRegVal}}\right)
+
\symCount_{\symRegRange+1} \log\left(1-e^{-\frac{\symPoissonRate}{\symNumReg 2^{\symRegRange}}}\right)
\end{equation}
and
\begin{equation}
\frac{d}{d\symPoissonRate}\log \mathcal{\symLikelihood}(\symPoissonRate\vert\vec{\symRegVal}) 
=
-\frac{1}{\symPoissonRate}\left(
\frac{\symPoissonRate}{\symNumReg}\sum_{\symRegVal=0}^\symRegRange \frac{\symCount_\symRegVal}{2^\symRegVal}+
\sum_{\symRegVal=1}^\symRegRange \symCount_\symRegVal\frac{\frac{\symPoissonRate}{\symNumReg 2^\symRegVal}}{1-e^{\frac{\symPoissonRate}{\symNumReg 2^\symRegVal}}}
+\symCount_{\symRegRange+1}\frac{\frac{\symPoissonRate}{\symNumReg 2^\symRegRange}}{1-e^{\frac{\symPoissonRate}{\symNumReg 2^\symRegRange}}}
\right).
\end{equation}
As a consequence, the maximum-likelihood estimate for the Poisson parameter is given by 
\begin{equation}
\symPoissonRateEstimate = \symNumReg\symXEstimate,
\end{equation}
if $\symXEstimate$ is the root of the function
\begin{equation}
\label{equ:funcdef}
\symFunc(\symX)
:=
\symX\sum_{\symRegVal=0}^\symRegRange \frac{\symCount_\symRegVal}{2^\symRegVal}+
\sum_{\symRegVal=1}^\symRegRange \symCount_\symRegVal\frac{\frac{\symX}{2^\symRegVal}}{1-e^{\frac{\symX}{2^\symRegVal}}}
+\symCount_{\symRegRange+1}\frac{\frac{\symX}{2^\symRegRange}}{1-e^{\frac{\symX}{2^\symRegRange}}}.
\end{equation}
which can also be written as
\begin{equation}
\label{equ:func}
\symFunc(\symX)
:=
\symX\sum_{\symRegVal=0}^\symRegRange \frac{\symCount_\symRegVal}{2^\symRegVal}+
\sum_{\symRegVal=1}^\symRegRange \symCount_\symRegVal\symHelper\left(\frac{\symX}{2^\symRegVal}\right)
+
\symCount_{\symRegRange+1}\symHelper\left(\frac{\symX}{2^\symRegRange}\right)
-
\left(\symNumReg-\symCount_0\right),
\end{equation}
if, the function $\symHelper(\symX)$ is defined as
\begin{equation}
\label{equ:helper}
\symHelper(\symX):=1-\frac{\symX}{e^{\symX}-1}.
\end{equation}
$\symHelper(\symX)$ is strictly increasing and concave as can be seen in \autoref{fig:helper_function}. For non-negative values $\symX$ the function ranges from $\symHelper(0)=0$ to $\symHelper(\symX\rightarrow \infty)=1$.
\begin{figure}
\centering
\includesvg[width=0.45\textwidth]{helper}
\caption{The function $\symHelper(\symX)$.}
\label{fig:helper_function}
\end{figure}
Since the function $\symFunc(x)$ is also strictly increasing, it is obvious that there exists a unique root $\symXEstimate$ for which $\symFunc(\symXEstimate)=0$. The function is non-positive at 0 since $\symFunc(0)=\symCount_0-\symNumReg\leq 0$ and, in case $\symCount_{\symRegRange+1}<\symNumReg$ which implies $\sum_{\symRegVal=0}^\symRegRange \frac{\symCount_\symRegVal}{2^\symRegVal}>0$, the function is at least linearly increasing. $\symCount_{\symRegRange}=\symNumReg$ corresponds to the case with all registers equal to the maximum value $(\symRegRange+1)$, for which the maximum likelihood estimate would be infinity.

It is easy to see that the $\symPoissonRateEstimate$ remains equal or is larger after updating the HyperLogLog data structure as defined by \algoref{alg:insert}. An update potentially changes the cardinality vector $(\symCount_0,\ldots,\symCount_{\symRegRange+1})$ to $(\symCount_0,\ldots,\symCount_\symIndexI-1,\ldots,\symCount_\symIndexJ+1,\ldots,\symCount_{\symRegRange+1})$ where $\symIndexI < \symIndexJ$. Writing \eqref{equ:func} as 
\begin{multline}
\symFunc(\symX)
:=
\symCount_0 \symX
+
\symCount_1 \left(
\symHelper\left(\frac{x}{2^1}\right)
+\frac{\symX}{2^1}-1
\right)
+
\symCount_2 \left(
\symHelper\left(\frac{x}{2^2}\right)
+\frac{\symX}{2^2}-1
\right)
+
\ldots
\\
\dots
+
\symCount_\symRegRange
\left(
\symHelper\left(\frac{\symX}{2^\symRegRange}\right)
+\frac{\symX}{2^\symRegRange}-1
\right)
+
\symCount_{\symRegRange+1}
\left(
\symHelper\left(\frac{\symX}{2^{\symRegRange}}\right)
-1
\right).
\end{multline}
shows that the coefficient of $\symCount_\symIndexI$ is larger than the coefficient of $\symCount_\symIndexJ$ in case $\symIndexI < \symIndexJ$. Keeping $\symX$ fixed during an update decreases $\symFunc(\symX)$. As a consequence, since $\symFunc(\symX)$ is increasing the new root and hence the estimate must be larger than before the update.

@@

\subsection{The new cardinality estimation algorithm}
In order to get a fast cardinality estimation algorithm, it is crucial to minimize the evaluation costs for \eqref{equ:funcprime}. A couple of optimizations allow significant reduction of the computational effort:
\begin{itemize}
\item Only a fraction of all count values $\symCount_\symRegVal$ is non-zero. If we denote $\symRegVal_\text{min}:=\min\lbrace \symRegVal\vert\symCount_\symRegVal>0\rbrace$ and $\symRegVal_\text{max}:=\max\lbrace \symRegVal\vert\symCount_\symRegVal>0\rbrace$,  it is sufficient to loop over all indices in the range $[\symRegVal_\text{min}, \symRegVal_\text{max}]$.
\item It is sufficient to calculate the coefficient of the linear term $\sum_{\symRegVal=0}^\symRegRange \frac{\symCount_\symRegVal}{2^\symRegVal}$ once and reuse it for any further iterations of the secant method.
\item Many programming languages allow the efficient multiplication and division by any integral power of 2 using special functions, such as \texttt{ldexp} in C/C++ or \texttt{scalb} in Java.
\item The function $\symHelper(\symX)$ only needs to be evaluated at values $\lbrace\frac{\symX}{2^{\symRegVal'_\text{max}}},\frac{\symX}{2^{\symRegVal'_\text{max}-1}}\ldots,\frac{\symX}{2^{\symRegVal_\text{min}}}\rbrace$ where $\symRegVal'_\text{max} := \min(\symRegVal_\text{max}, \symRegRange)$. This series corresponds to geometric series with ratio 2. A straightforward calculation using \eqref{equ:helper} is very expensive because of the exponential function. However, if we know $\symHelper\left(\frac{\symX}{2^{\symRegVal'_\text{max}}}\right)$ all other required function values can be easily obtained using the identity
\begin{equation}
\label{equ:helper_recursion1}
\symHelper\left(2\symX\right) = \frac{\symX+2\symHelper\left(\symX\right)\left(1-\symHelper\left(\symX\right)\right)}{\symX+2\left(1-\symHelper\left(\symX\right)\right)}
\end{equation}
or
\begin{equation}
\label{equ:helper_recursion2}
\symHelper\left(\frac{\symX}{2^{\symRegVal}}\right) = \frac{\frac{\symX}{2^{\symRegVal+2}}+\symHelper\left(\frac{\symX}{2^{\symRegVal+1}}\right)\left(1-\symHelper\left(\frac{\symX}{2^{\symRegVal+1}}\right)\right)}{\frac{\symX}{2^{\symRegVal+2}}+\left(1-\symHelper\left(\frac{\symX}{2^{\symRegVal+1}}\right)\right)}
\end{equation}
Note, this recursive formula is stable in a sense that the relative error of $\symHelper\left(2\symX\right)$ is smaller than that of $\symHelper\left(\symX\right)$ as shown in \autoref{app:helper_stable}.

>>>>

The number of bits of the consumed hash value $\symPrecision+\symRegRange$ limit the maximum cardinality that can be tracked. Obviously, if the cardinality reaches values in the order of $2^{\symPrecision+\symRegRange}$, hash collisions will become more apparent, which will reduce estimation accuracy drastically.

@@

For large cardinalities in the order of $2^{\symPrecision+\symRegRange}$, for which a lot of registers are already in a saturated state meaning that they have reached the maximum possible value $\symRegRange + 1$, the \emph{raw} estimate underestimates the cardinalities. For the 32-bit hash value case $(\symPrecision+\symRegRange=32)$ which was considered in \cite{Flajolet2007}, following correction formula was proposed to take these saturated registers into account
\begin{equation}
\label{equ:large_range_estimate}
\symCardinalityEstimate_\text{large}
=
-2^{32}\log\left(1-\frac{\symCardinalityOriginalEstimate}{2^{32}}\right).
\end{equation}

The original estimation algorithm as presented in \cite{Flajolet2007} including small and large range corrections is summarized by \algoref{alg:estimate_original}. 
\begin{algorithm}
\caption{Original procedure for estimating the cardinality from a  HyperLogLog data structure using 32-bit hash values ($\symPrecision+\symRegRange = 32$) for insertion of data items \cite{Flajolet2007}.}
\label{alg:estimate_original}
\begin{algorithmic}
\Function {EstimateCardinality}{$\vec{\symRegVal}$}
\State $\symNumReg \gets \dim \vec{\symRegVal}$
\State $\symCardinalityOriginalEstimate = \symAlpha_\symNumReg \symNumReg^2\left(\sum_{\symIndexJ=1}^\symNumReg 2^{-\symRegVal_\symIndexJ}\right)^{-1}$
\Comment \emph{raw} estimate \eqref{equ:intermediate_estimate}
\If{$\symCardinalityOriginalEstimate\leq \frac{5}{2}\symNumReg$}
\State $\symCount_0 = \left|\{\symIndexJ |\symRegVal_\symIndexJ=0\}\right|$
\If{$\symCount_0\neq0$}
\State \Return $\symNumReg \log(\symNumReg / \symCount_0)$ \Comment small range correction \eqref{equ:small_range_estimate}
\Else
\State \Return $\symCardinalityOriginalEstimate$
\EndIf
\ElsIf{$\symCardinalityOriginalEstimate\leq \frac{1}{30}2^{32}$}
\State \Return $\symCardinalityOriginalEstimate$
\Else
\State\Return $-2^{32}\log(1-\symCardinalityOriginalEstimate/2^{32})$\Comment large range correction \eqref{equ:large_range_estimate}
\EndIf
\EndFunction
\end{algorithmic}
\end{algorithm}
The relative estimation error for the original method is shown in \autoref{fig:original_estimate}. Unfortunatley, as can be clearly seen, the ranges where the estimation error is small for $\symCardinalityOriginalEstimate$ and $\symCardinalityEstimate_\text{small}$ do not overlap. Therefore, the estimation error is much larger near the transition region. To reduce the estimation error for cardinalities close to this region, it was proposed to correct $\symCardinalityOriginalEstimate$ for bias. The empirically determined bias correction data can be either stored as set of interpolation points \cite{Heule2013}, as lookup table \cite{Rhodes2015}, or as best-fitting polynomial \cite{Sanfilippo2014}.

@@

The large range correction is not satisfying either as it does not reduce the estimation error. Quite the contrary, it even increases the estimation error. However, instead of underestimating the cardinalities, they are now overestimated. A simple approach to avoid the large range correction at all is the use of larger hash values $(\symPrecision+\symRegRange>32)$ which shift the biased region of $\symCardinalityOriginalEstimate$ to higher cardinalities. However, in case $\symRegRange \geq 30$ 6-bit registers as used in \cite{Heule2013} are required to allow representing all values in the range $[0, \symRegRange+1]$. If 64-bit hash values are used as proposed in \cite{Heule2013} $(\symPrecision+\symRegRange=64)$, a correction for large cardinalities is not required. Cardinalities in the order of $2^{64}$ for which saturated register become a problem never occur in practice.

@@

\section{Cardinality estimation}
\label{sec:cardinality_estimation}
In the following we describe a new cardinality estimation method that does not require special handling for small or large cardinalities and which also does not require any empirically determined constants or corrections. The method is based on the maximum likelihood method.

The maximum likelihood method was already applied to HyperLogLog sketches in \cite{Clifford2012}. However, there, a different element insertion algorithm was assumed that works with time complexity $\symBigO(\symNumReg)$, because each register calculates its own hash value from the inserted element and updates its current value accordingly. In comparison, \algoref{alg:insert} uses only a single hash value that defines which register needs to be updated. As a consequence, the register values are statistically dependent, which makes the joint probability distribution of all register values more complex.

In the following we use the Poisson approximation in order to simplify the model. Despite this approximation the maximum likelihood estimate will give us high quality estimates for the entire cardinality range as simulation results will show later. Furthermore, we will demonstrate that consequent application of the maximum likelihood method to the Poisson model directly reveals that the cardinality estimate needs to be roughly proportional to the harmonic mean for intermediate cardinality values. The history of the HyperLogLog algorithm shows that the \emph{raw} estimate \eqref{alg:estimate_original} was first found after several attempts using the geometric mean \cite{Flajolet2007, Durand2003}.

@@

Under the assumption of an ideal hash function, the probability mass function for the register values $\vec{\symRegVal}=(\symRegVal_1,\ldots,\symRegVal_\symNumReg)$ of a HyperLogLog sketch with parameters $\symPrecision$ and $\symRegRange$ is given by
\begin{equation}
\label{equ:multinomialProbabilityMass}
\symProbabilityMass(\vec{\symRegVal}\vert\symCardinality)
=
\sum_{\symCardinality_1+\ldots+\symCardinality_\symNumReg = \symCardinality} \binom{\symCardinality}{\symCardinality_1,\ldots\symCardinality_\symNumReg}
\frac{1}{\symNumReg^\symCardinality}\prod_{\symIndexJ=1}^\symNumReg \symRegProbability_{\symCardinality_\symIndexJ\symRegVal_\symIndexJ},
\end{equation}
where $\symCardinality$ is the cardinality. The $\symCardinality$ distinct elements are distributed over all $\symNumReg$ registers according to a multinomial distribution with equal probabilities. $\symRegProbability_{\symNu\symKappa}$ is the probability that a register is equal to $\symKappa$ after inserting $\symNu$ distinct elements
\begin{equation}
\symRegProbability_{\symNu\symKappa} 
:=
\begin{cases}
1 & \symNu=0 \wedge \symKappa = 0\\
0 & \symNu=0 \wedge 1\leq\symKappa\leq\symRegRange+1\\
0 & \symNu\geq1 \wedge \symKappa = 0\\
\left(1-\frac{1}{2^\symKappa}\right)^\symNu - \left(1-\frac{1}{2^{\symKappa-1}}\right)^\symNu & \symNu \geq 1 \wedge 1\leq\symKappa\leq\symRegRange\\
1 - \left(1-\frac{1}{2^{\symRegRange}}\right)^\symNu & \symNu\geq 1 \wedge \symKappa = \symRegRange +1 \\
\end{cases}
\end{equation}
Note that the order of register values $\symRegVal_1,\ldots,\symRegVal_\symNumReg$ is not important for the estimation of the cardinality. More formally, the multiset $\lbrace\symRegVal_1,\ldots,\symRegVal_\symNumReg\rbrace$ is a sufficient statistic for $\symCardinality$.
Since the values of the multiset are all in the range $[0, \symRegRange+1]$ the multiset can also be represented as $\lbrace\symRegVal_1,\ldots,\symRegVal_\symNumReg\rbrace = 0^{\symCount_0}1^{\symCount_1}\cdots\symRegRange^{\symCount_{\symRegRange}}(\symRegRange+1)^{\symCount_{\symRegRange+1}}$ where $\symCount_\symIndexJ$ is the multiplicity of value $\symIndexJ$. As a consequence, the multiplicity vector $\vec{\symCount} := (\symCount_0,\ldots,\symCount_{\symRegRange+1})$ is also a sufficient statistic for the cardinality. In addition, this vector also contains all the information about the HyperLogLog sketch that is required for  cardinality estimation: $\symRegRange=\dim(\vec{\symCount})-2$ and  $\symPrecision = \log_2 \left\Vert\vec{\symCount}\right\Vert_1$.

@@

As supported by simulations that will be presented later, this approximation works well for the entire range of cardinalities. In order to get an estimate for $\symPoissonRateEstimate$, maximum-likelihood estimation is used which works out quite well under the Poisson model. The log-likelihood and its derivative are given by
\begin{equation}
\label{equ:log_likelihood_single}
\log \mathcal{\symLikelihood}(\symPoissonRate\vert\vec{\symRegVal}) = 
-\frac{\symPoissonRate}{\symNumReg}\sum_{\symRegVal=0}^{\symRegRange}\frac{\symCount_\symRegVal}{2^\symRegVal}+ 
\sum_{\symRegVal=1}^{\symRegRange}\symCount_\symRegVal \log\left(1-e^{-\frac{\symPoissonRate}{\symNumReg 2^\symRegVal}}\right)
+
\symCount_{\symRegRange+1} \log\left(1-e^{-\frac{\symPoissonRate}{\symNumReg 2^{\symRegRange}}}\right)
\end{equation}
and
\begin{equation}
\frac{d}{d\symPoissonRate}\log \mathcal{\symLikelihood}(\symPoissonRate\vert\vec{\symRegVal}) 
=
-\frac{1}{\symPoissonRate}\left(
\frac{\symPoissonRate}{\symNumReg}\sum_{\symRegVal=0}^\symRegRange \frac{\symCount_\symRegVal}{2^\symRegVal}+
\sum_{\symRegVal=1}^\symRegRange \symCount_\symRegVal\frac{\frac{\symPoissonRate}{\symNumReg 2^\symRegVal}}{1-e^{\frac{\symPoissonRate}{\symNumReg 2^\symRegVal}}}
+\symCount_{\symRegRange+1}\frac{\frac{\symPoissonRate}{\symNumReg 2^\symRegRange}}{1-e^{\frac{\symPoissonRate}{\symNumReg 2^\symRegRange}}}
\right).
\end{equation}
As a consequence, the maximum-likelihood estimate for the Poisson parameter is given by 
\begin{equation}
\symPoissonRateEstimate = \symNumReg\symXEstimate,
\end{equation}
if $\symXEstimate$ is the root of the function
\begin{equation}
\label{equ:funcdef}
\symFunc(\symX)
:=
\symX\sum_{\symRegVal=0}^\symRegRange \frac{\symCount_\symRegVal}{2^\symRegVal}+
\sum_{\symRegVal=1}^\symRegRange \symCount_\symRegVal\frac{\frac{\symX}{2^\symRegVal}}{1-e^{\frac{\symX}{2^\symRegVal}}}
+\symCount_{\symRegRange+1}\frac{\frac{\symX}{2^\symRegRange}}{1-e^{\frac{\symX}{2^\symRegRange}}}.
\end{equation}
This function can also be written as
\begin{equation}
\label{equ:func}
\symFunc(\symX)
:=
\symX\sum_{\symRegVal=0}^\symRegRange \frac{\symCount_\symRegVal}{2^\symRegVal}+
\sum_{\symRegVal=1}^\symRegRange \symCount_\symRegVal\symHelper\left(\frac{\symX}{2^\symRegVal}\right)
+
\symCount_{\symRegRange+1}\symHelper\left(\frac{\symX}{2^\symRegRange}\right)
-
\left(\symNumReg-\symCount_0\right),
\end{equation}
where the function $\symHelper(\symX)$ is defined as
\begin{equation}
\label{equ:helper}
\symHelper(\symX):=1-\frac{\symX}{e^{\symX}-1}.
\end{equation}
$\symHelper(\symX)$ is strictly increasing and concave as can be seen in \autoref{fig:helper_function}. For non-negative values $\symX$ the function ranges from $\symHelper(0)=0$ to $\symHelper(\symX\rightarrow \infty)=1$.
\begin{figure}
\centering
\includesvg[width=0.45\textwidth]{helper}
\caption{The function $\symHelper(\symX)$.}
\label{fig:helper_function}
\end{figure}
Since the function $\symFunc(x)$ is also strictly increasing, it is obvious that there exists a unique root $\symXEstimate$ for which $\symFunc(\symXEstimate)=0$. The function is non-positive at 0 since $\symFunc(0)=\symCount_0-\symNumReg\leq 0$ and, in case $\symCount_{\symRegRange+1}<\symNumReg$ which implies $\sum_{\symRegVal=0}^\symRegRange \frac{\symCount_\symRegVal}{2^\symRegVal}>0$, the function is at least linearly increasing. $\symCount_{\symRegRange}=\symNumReg$ corresponds to the case with all registers equal to the maximum value $(\symRegRange+1)$, for which the maximum likelihood estimate would be infinity.

It is easy to see that the estimate $\symPoissonRateEstimate$ remains equal or becomes larger, when inserting an element into the HyperLogLog sketch following \algoref{alg:insert}. An update potentially changes the multiplicity vector $(\symCount_0,\ldots,\symCount_{\symRegRange+1})$ to $(\symCount_0,\ldots,\symCount_\symIndexI-1,\ldots,\symCount_\symIndexJ+1,\ldots,\symCount_{\symRegRange+1})$ where $\symIndexI < \symIndexJ$. Writing \eqref{equ:func} as 
\begin{multline}
\symFunc(\symX)
:=
\symCount_0 \symX
+
\symCount_1 \left(
\symHelper\left(\frac{x}{2^1}\right)
+\frac{\symX}{2^1}-1
\right)
+
\symCount_2 \left(
\symHelper\left(\frac{x}{2^2}\right)
+\frac{\symX}{2^2}-1
\right)
+
\ldots
\\
\dots
+
\symCount_\symRegRange
\left(
\symHelper\left(\frac{\symX}{2^\symRegRange}\right)
+\frac{\symX}{2^\symRegRange}-1
\right)
+
\symCount_{\symRegRange+1}
\left(
\symHelper\left(\frac{\symX}{2^{\symRegRange}}\right)
-1
\right).
\end{multline}
shows that the coefficient of $\symCount_\symIndexI$ is larger than the coefficient of $\symCount_\symIndexJ$ in case $\symIndexI < \symIndexJ$. Keeping $\symX$ fixed during an update decreases $\symFunc(\symX)$. As a consequence, since $\symFunc(\symX)$ is increasing, the new root and hence the estimate must be larger than before the update.

@@

\subsection{The new cardinality estimation algorithm}
In order to get a fast cardinality estimation algorithm, it is crucial to minimize the evaluation costs for \eqref{equ:funcprime}. A couple of optimizations allow significant reduction of the computational effort:
\begin{itemize}
\item Only a fraction of all count values $\symCount_\symRegVal$ is non-zero. If we denote $\symRegVal_\text{min}:=\min\lbrace \symRegVal\vert\symCount_\symRegVal>0\rbrace$ and $\symRegVal_\text{max}:=\max\lbrace \symRegVal\vert\symCount_\symRegVal>0\rbrace$,  it is sufficient to loop over all indices in the range $[\symRegVal_\text{min}, \symRegVal_\text{max}]$.
\item It is sufficient to calculate the coefficient of the linear term $\sum_{\symRegVal=0}^\symRegRange \frac{\symCount_\symRegVal}{2^\symRegVal}$ once and reuse it for any further iterations of the secant method.
\item Many programming languages allow the efficient multiplication and division by any integral power of 2 using special functions, such as \texttt{ldexp} in C/C++ or \texttt{scalb} in Java.
\item The function $\symHelper(\symX)$ only needs to be evaluated at values $\lbrace\frac{\symX}{2^{\symRegVal'_\text{max}}},\frac{\symX}{2^{\symRegVal'_\text{max}-1}}\ldots,\frac{\symX}{2^{\symRegVal_\text{min}}}\rbrace$ where $\symRegVal'_\text{max} := \min(\symRegVal_\text{max}, \symRegRange)$. This series corresponds to a geometric series with ratio 2. A straightforward calculation using \eqref{equ:helper} is very expensive because of the exponential function. However, if we know $\symHelper\left(\frac{\symX}{2^{\symRegVal'_\text{max}}}\right)$ all other required function values can be easily obtained using the identity
\begin{equation}
\label{equ:helper_recursion1}
\symHelper\left(2\symX\right) = \frac{\symX+2\symHelper\left(\symX\right)\left(1-\symHelper\left(\symX\right)\right)}{\symX+2\left(1-\symHelper\left(\symX\right)\right)}
\end{equation}
or
\begin{equation}
\label{equ:helper_recursion2}
\symHelper\left(\frac{\symX}{2^{\symRegVal}}\right) = \frac{\frac{\symX}{2^{\symRegVal+2}}+\symHelper\left(\frac{\symX}{2^{\symRegVal+1}}\right)\left(1-\symHelper\left(\frac{\symX}{2^{\symRegVal+1}}\right)\right)}{\frac{\symX}{2^{\symRegVal+2}}+\left(1-\symHelper\left(\frac{\symX}{2^{\symRegVal+1}}\right)\right)}
\end{equation}
Note, this recursive formula is stable in a sense that the relative error of $\symHelper\left(2\symX\right)$ is smaller than that of $\symHelper\left(\symX\right)$ as shown in \autoref{app:helper_stable}.
